---
documentclass: jss
author:
  - name: James Hollway
    affiliation: |
      | Graduate Institute of
      | International and Development Studies
    # use this syntax to add text on several lines
    address: |
      | Chemin Eugène-Rigot 2A 
      | PO Box 1672 
      | 1211 Geneva 1 
      | Switzerland
    email: \email{james.hollway@graduateinstitute.ch}
    url: http://jameshollway.com
  - name: Henrique Sposito
    affiliation: |
      | Graduate Institute of
      | International and Development Studies
title:
  formatted: "Working with Unspecified, Approximate, Uncertain, Sets and Ranges of Dates with \\pkg{messydates}"
  # If you use tex in the formatted title, also supply version without
  plain: "Working with Unspecified, Approximate, Uncertain, Sets and Ranges of Dates with messydates"
  # For running headers, if needed
  short:     "\\pkg{messydates}: An R package for ISO's Extended Date/Time Format"
abstract: >
  This paper presents the \pkg{messydates} package for R, which facilitates working with 'messy' dates. 
  Messy dates are common when studying historical and sometimes even current phenomena, 
  and can create various technical problems for the data analyst. 
  The paper highlights these problems and offers practical advice on how to solve them using \pkg{messydates}. 
  The paper also introduces a conceptual framework for resolving messydates into more familiar date classes in R ready for analysis.
keywords:
  # at least one keyword must be supplied
  formatted: [dates, ISO, "\\proglang{R}"]
  plain:     [dates, ISO, R]
bib_file: 'references.bib'
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: references.bib
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
library(messydates)
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(anytime))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(dplyr))
```

# Introduction

Dates are often messy.
Whether historical (or ancient), future, or even recent,
we often only know approximately when an event occurred,
that it happened within a particular period,
an unreliable source means a date should be flagged as uncertain,
or sources offer multiple, competing dates.

\pkg{messydates} implements for R the Extended Date/Time Format (EDTF) annotations
set by the International Organization for Standardization (ISO) 
outlined in [ISO 8601-2_2019(E)](https://www.iso.org/standard/70908.html).
The standardised extended format allow for unambiguous interpretation of dates
and guarantee interoperability.
These include notation for:

-   unspecified date(component)s, e.g. `2012-XX-01` for the first of some unknown month in 2012 or `2012-01` for some unknown day in January 2012
-   approximate date(component)s, e.g. `2012-01-12~` for approximately the 12th of January 2012
-   uncertain date(component)s, e.g. `2012-01-12?` where this data point is based on an unreliable source
-   sets of dates, e.g. `{2012-01-01,2012-01-12}` where the date can be either 1 January 2012 and 12 January 2012
-   ranges of dates, e.g. `2012-01-01..2012-01-12` for all dates between the 1 January 2012 and 12 January 2012 inclusive

\pkg{messydates} contains a set of tools for constructing and coercing into
and from the `mdate` class.
This date class allows regular dates to be annotated to express unspecified,
approximate or uncertain date components, date ranges, and sets of dates.

The package includes functions for expanding sets
or ranges of dates into all dates consistent with how the date or set of dates
are specified or annotated.
Methods are also offered that can be used to make explicit how researchers
convert date imprecision into precise dates for analysis,
such as getting the `min()`, `max()`, or even a `random()` date
from among the dates in a set or range of dates.
This greatly facilitates research transparency and robustness checks
as we will demonstrate below.

## Motivation

As researchers, we often recognize this messiness but are required to
force non-existent precision on data so we can proceed with analysis.
For example, if we only know something happened in a given month or year,
we might just opt for the start of that month (e.g. `2021-07-01`) or year (`2021-01-01`), assuming that to err on the earlier (or later) side is a justifiable bias.
However, this can create issues for inference in which sequence or timing is important.
The goal of \pkg{messydates} is to help with this problem by retaining and
working with various kinds of date imprecision.

## Relationship to other packages

\pkg{messydates} offers a new date class,
but one that comes with methods for converting from and into `base`
date classes such as `Date`, `POSIXct`, and `POSIXlt`.
It is thus fully compatible with packages such as \pkg{lubridate} [@grolemundDatesTimesMade2011]
and \pkg{anytime} [@eddelbuettelAnytimeEasierDate2019].
\pkg{messydates} is, therefore, compatible (perhaps with an additional coercion step)
with all contemporary R packages for analysis.

# \proglang{R} code {#r-code short-title="R code"}

## A new class

\pkg{messydates} contains a set of tools for constructing and
coercing into and from the `mdate` class.
This date class implements ISO 8601-2:2019(E) and allows regular dates to be
annotated to express unspecified date components,
approximate or uncertain date components, date ranges, and sets of dates.
The function `as_messydate()` handles the coercion to the `mdate` class.

```{r comparison, warning=FALSE}
tibble::tribble(~Example, ~Date,
                "A normal date", as.character(Sys.Date()),
                "A future date", "2599-12-31",
                "A written date", "This is the first day of February, two thousand and twenty-one",
                "A historical date", "476",
                "An era date", "33 BC",
                "An approximate date", "2012-01-12~",
                "An uncertain date", "2001-01-01?",
                "An unspecified date", "2012-01",
                "A censored date", "..2012-01-12",
                "A range of dates", "2019-11-01:2020-01-01",
                "A set of dates", "2021-5-26, 2021-11-19, 2021-12-4") %>%
  dplyr::mutate(base = as.Date(Date),
                lubridate = lubridate::as_date(Date),
                anytime = anytime::anydate(Date),
                messydates = messydates::as_messydate(Date))
```

## Annotate

Some datasets have, for example, an arbitrary cut off point for start and end points, 
but these are often coded as precise dates when they are not 
the real start or end dates.
\pkg{messydates} helps annotate uncertainty and approximation to dates.
Inaccurate start or end dates can be represented by an affix
indicating "on or before", if used as a prefix (e.g. `..1816-01-01`),
or indicating "on or after", if used as a suffix (e.g. `2016-12-31..`).
Approximate dates are indicated by adding a `~` to year, month, or day components,
as well as groups of components or whole dates to estimate values that
are possibly correct (e.g. `2003-03-03~`).
Day, month, or year, uncertainty can be indicated by adding a `?` to a
possibly dubious date (e.g. `1916-10-10?`) or date component (e.g. `1916-?10-10`).

```{r annotate, warning=FALSE}
tibble::tibble(Beg = as_messydate(c("1816-01-01", "1916-01-01", "2016-01-01")),
               End = as_messydate(c("1816-12-31", "1916-12-31", "2016-12-31"))) %>% 
  dplyr::mutate(on_or_before = ifelse(Beg <= "1816-01-01", on_or_before(Beg), Beg),
                on_or_after = ifelse(End >= "2016-01-01", on_or_after(End), End),
                as_approximate = ifelse(End >= "2016-01-01", on_or_after(End), End),
                as_uncertain = ifelse(End == "1916-12-31", as_uncertain(End), End))
```

## Expand

The `expand()´ function transform date ranges, sets of dates, and unspecified or approximate dates (annotated with '..', '{ , }', 'XX' or '\~') into lists of dates.
As these dates may refer to several possible dates,
the function "opens" these values to include all the possible dates implied.

```{r expand, warning=FALSE}
dates_expand <- as_messydate(c("2001-01-01", "2001-01", "2001-01-01..2001-02-01",
                               "{2001-01-01,2001-02-01}", "2001-XX-01"))
expand(dates_expand)
```

## Contract

The `contract()` function operates as the opposite of `expand()`.
It contracts a list of dates into their abbreviated annotations.

```{r contract, warning=FALSE, message=FALSE}
tibble::tibble('Original Dates' = dates_expand,
               'Contracted Dates' = contract(expand(dates_expand)))
```

## Coerce from messydates

Coercion functions coerce objects of `mdate` class objects to common date classes
such as `Date`, `POSIXct`, and `POSIXlt`.
Since `mdate` objects can hold multiple individual dates,
an additional function must be passed as an argument so
that multiple dates are "resolved" into a single date.
For example, one might wish to use the earliest possible date
in a range, or set, of expanded dates (`min`), or the latest possible date (`max`),
or some notion of a central tendency (`mean`, `median`, or `modal`),
or even a `random` selection from amongst the candidate dates.

```{r coerce, warning=FALSE, message=FALSE}
tibble::tibble(min = as.Date(dates_expand, min),
               max = as.Date(dates_expand, max),
               median = as.Date(dates_expand, median),
               mean = as.Date(dates_expand, mean),
               modal = as.Date(dates_expand, modal),
               random = as.Date(dates_expand, random))
```

## Additional functionality 

Several other functions are also offered in the \pkg{messydates} package.For example, one can run various logical tests for checking `mdate` objects:

-   `is_messydate()` tests whether the object inherits the `mdate` class
-   `is_intersecting()` tests whether there is any intersection between two `mdate` objects
-   `is_element()` similarly tests whether an `mdate` can be found within an `mdate` range or set
-   `is_similar()` tests whether two `mdate` contain similar components
-   `is_precise()` tests for whether `mdate` is precise

```{r logical, warning=FALSE}
is_messydate(as_messydate("2001-01-01"))
is_messydate(as.Date("2001-01-01"))
is_intersecting(as_messydate("2001-01"), as_messydate("2001-01-01..2001-02-22"))
is_intersecting(as_messydate("2001-01"), as_messydate("2001-02-01..2001-02-22"))
is_element(as_messydate("2001-01-01"), as_messydate("2001-01"))
is_element(as_messydate("2001-01-01"), as_messydate("2001-02"))
is_similar(as_messydate("2001-06-02"), as_messydate("2001-02-06"))
is_similar(as_messydate("2001-06-22"), as_messydate("2001-02-06"))
is_precise(as_messydate("2001-06-02"))
is_precise(as_messydate("2001-02"))
```

Additionally, one can perform intersection (`md_intersect()`) and union (`md_union()`) on, inter alia, messy date class objects.
Or perform a 'join' that retains all elements, even if duplicated, with `md_multiset`.

```{r set, warning=FALSE}
md_intersect(as_messydate("2001-01-01..2001-01-20"), as_messydate("2001-01"))
md_union(as_messydate("2001-01-01..2001-01-20"), as_messydate("2001-01"))
md_multiset(as_messydate("2001-01-01..2001-01-20"), as_messydate("2001-01"))
```

Some arithmetic operations are available for messydates.
For instance, one can add, or subtract, a day or one year to one, or all, `mdate` objects in a vector.

```{r operate, warning=FALSE, message=FALSE}
tibble::tibble(date = dates_expand,
               add = dates_expand + 1,
               subtract = dates_expand - "1 year")
```

## Case Study - 2001 Battles

Dates, even for some recent events, can be messy.
Take the dates of battles in 2001 according to 
[Wikipedia](https://en.wikipedia.org/wiki/List_of_battles_in_the_21st_century)
included in \pkg{messydates}.
The dates of these battles are often approximate
(i.e. the day in which a battle started or ended is unknown)
or come from unreliable sources (i.e. the date might not be trustworthy).

```{r data, warning=FALSE, message=FALSE}
battles <- messydates::battles
battles
```

\pkg{messydates} facilitates working with these dates as we can, for example,
check date precision, get the median values, and find the longest battle in 2001.

```{r working with dates, warning=FALSE, message=FALSE}
messydates::is_precise(battles$Date)
as.Date(battles$Date, median)
as.numeric(as.Date(battles$Date, max) - as.Date(battles$Date, min))
```

Getting the timing right can be important for researchers.
This is especially true if researchers are looking to generate robust inferences.
Until now, when faced with date imprecision, researchers usually have to choose
between making arbitrary choices (e.g. adding "-01-01" to non-precise dates)
or work with imprecise dates (e.g. year only).
Either choice may lead to biased results.
Assume we are interested in the relationship between the United States (US)
being a party to a conflict and the duration of the conflict in 2001.
We hypothesize that conflicts involving the US have
a shorter duration because the US military capabilities.
Using \pkg{messydates} we create two different variables representing conflict time from the 2001 battles data to be our dependent variables;
one variable with arbitrary cut off points and the other variable
with random values for uncertain or approximate dates.
As our independent variable, we create a binary variable for whether the US
was involved in the conflict.
As a control, we code the number of actors in the conflict.
With these variables we run two simple linear regression models.

```{r lm, warning=FALSE, message=FALSE}
set.seed(1301)
battles <- battles %>%
  mutate(arbitrary = as.numeric(as.Date(Date, max) - as.Date(Date, min)),
         random = ifelse(is_uncertain(Date)|is_approximate(Date),
                         abs(as.Date(Date, random) - as.Date(Date, random)),
                         arbitrary),
         US_party = ifelse(grepl("US", Parties), 1, 0),
         n_actors = c(2, 2, 2, 2, 2, 3, 2, 4, 4, 4, 3, 3, 4, 4, 5, 4, 3, 3, 2, 2))
lm(arbitrary ~ US_party + n_actors, battles)
lm(random ~ US_party + n_actors, battles)
```

Notice how the regression coefficients change in the two models.
Although not statistically significant, the coefficient for US being a party
in a conflict goes from being positive,
when calculated using arbitrary cut off dates, 
to being negative, when calculated using random dates.
In this case, setting arbitrary cut off points to dates introduces highly
influential outliers (see Cook's distance below).

```{r outliers, warning=FALSE, message=FALSE}
plot(lm(arbitrary ~ US_party + n_actors, battles), which = 4)
```

Hence, it is hard to say whether there is a relationship between the US being
an actor involved in one of the battles in 2001 and its duration.

# Acknowledgements

We would like to thank the Swiss National Science Foundation. 
This work was supported by grant number 188976.

# References
