---
title: "Working with messy dates"
author: "Henrique Sposito"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to messydates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  purl = NOT_CRAN,
  eval = NOT_CRAN
)
```

## Why `{messydates}`?

Dates are often messy.
Whether historical (or ancient), future, or even recent,
we often only know approximately when an event occurred,
that it happened within a particular period,
an unreliable source means a date should be flagged as uncertain,
or sources offer multiple, competing dates.

As researchers, we often recognise this messiness
but are forced to force non-existent precision on data so we can proceed with analysis.
For example, if we only know something happened in a given month or year,
we might just opt for the start of that month (e.g. 2021-07-01) or year (2021-01-01),
assuming that to err on the earlier (or later) side is a justifiable bias.
However, this can create issues for inference in which sequence or timing is important.
The goal of `{messydates}` is to help with this problem by retaining and working
with various kinds of date imprecision.

`{messydates}` contains a set of tools for constructing and coercing into and from the messydt class. 
This date class implements ISO 8601-2:2019(E) and allows regular dates to be annotated 
to express unspecified date components, approximate or uncertain date components, 
date ranges, and sets of dates. 
The function `as_messydate()` handles the coercion to messydt class.

```{r comparison, warning=FALSE}
library(messydates)
library(lubridate)
library(tibble)
library(dplyr)
dates_comparison <- tibble::tribble(~Example, ~OriginalDate,
                                    "A normal date", "2010-01-01",
                                    "A historical date", "1291-08-01",
                                    "A very historical date", "476",
                                    "A really historical date", "33 BC",
                                    "A clearly future date", "9999-12-31",
                                    "A not so clearly future date", "2599-12-31",
                                    "A range of dates", "2019-11-01:2020-01-01",
                                    "An uncertain date", "2001-01-01?",
                                    "A set of dates", "2021-5-26, 2021-6-10, 2021-11-19, 2021-12-4")
dates_comparison %>% dplyr::mutate(base = as.Date(OriginalDate),
                                   lubridate = suppressWarnings(lubridate::as_date(OriginalDate)),
                                   messydates = messydates::as_messydate(OriginalDate)) %>%
  print()
```

## Annotate

Some datasets have, for example, an arbitrary cut off point for start and end points, 
but these are often coded as precise dates when they are not necessarily the real start or end dates.
The annotate functions helps annotate uncertainty and approximation to dates.
Inaccurate start or end dates can be represented by an affix
indicating "on or before", if used as a prefix (e.g. ..1816-01-01),
or indicating "on or after", if used as a suffix (e.g. 2016-12-31..).
Approximate dates are indicated by adding a `~` to year,
month, or day components, as well as groups of components or whole dates
to estimate values that are possibly correct (e.g. 2003-03-03~).
Day, month, or year, uncertainty can be indicated by adding a `?`
to a possibly dubious date (e.g. 1916-10-10?) or date
component (e.g. 1916-?10-10).

```{r annotate, warning=FALSE}
dates_annotate <- tibble::tibble(Beg = as_messydate(c("1816-01-01", "1916-01-01", "2016-01-01")),
                                 End = as_messydate(c("1816-12-31", "1916-12-31", "2016-12-31")))
dplyr::mutate(dates_annotate, Beg = ifelse(Beg <= "1816-01-01", on_or_before(Beg), Beg))
dplyr::mutate(dates_annotate, End = ifelse(End >= "2016-01-01", on_or_after(End), End))
dplyr::mutate(dates_annotate, Beg = ifelse(Beg == "1916-01-01", as_approximate(Beg), Beg))
dplyr::mutate(dates_annotate, End = ifelse(End == "1916-12-31", as_uncertain(End), End))
```

## Expand

Expand functions transform date ranges, sets of dates, and unspecified or 
approximate dates (annotated with '..', '{ , }', 'XX' or '~') into lists of dates.
As these dates may refer to several possible dates,
the function "opens" these values to include all the possible dates implied.

```{r expand, warning=FALSE}
dates_expand <- as_messydate(c("2008-03-25", "2001-01?", "2001",
                               "2001-01-01..2001-02-02", "{2001-01-01,2001-02-02}",
                               "2008-XX-31", "28 BC"))
expand(dates_expand)
```

## Contract

The `contract()` function operates as the opposite of `expand()`.
It contracts a list of dates into the abbreviated annotation of messydates.

```{r contract, warning=FALSE}
tibble::tibble(contract = contract(expand(dates_expand)))
```

## Coerce from messydates

Coercion functions coerce objects of `messydt` class to 
common date classes such as `Date`, `POSIXct`, and `POSIXlt`.
Since `messydt` objects can hold multiple individual dates,
an additional function must be passed as an argument so that multiple dates
are "resolved" into a single date.

For example, one might wish to use the earliest possible date
in any ranges of dates (`min`), the latest possible date (`max`),
some notion of a central tendency (`mean`, `median`, or `modal`),
or even a `random` selection from amongst the candidate dates.

These functions are particularly useful for use with existing methods and models,
especially for checking the robustness of results.

```{r coerce, warning=FALSE}
tibble::tibble(min = as.Date(dates_expand, min),
               max = as.Date(dates_expand, max),
               median = as.Date(dates_expand, median),
               mean = as.Date(dates_expand, mean),
               modal = as.Date(dates_expand, modal),
               random = as.Date(dates_expand, random))
```

## Additional functionality

Several other functions are also offered in the `{messydates}` package.

For example, one can check various logical tests for messy date objects.
`is_messydate()` tests whether the object inherits the `messydt` class.
`is_intersecting()` tests whether there is any intersection between two messy dates.
`is_element()` similarly tests whether a messy date can be found within a messy date range or set.
`is_similar()` tests whether two dates contain similar components.

```{r logical, warning=FALSE}
is_messydate(as_messydate("2012-01-01"))
is_messydate(as.Date("2012-01-01"))
is_intersecting(as_messydate("2012-01"), as_messydate("2012-01-01..2012-02-22"))
is_intersecting(as_messydate("2012-01"), as_messydate("2012-02-01..2012-02-22"))
is_element(as_messydate("2012-01-01"), as_messydate("2012-01"))
is_element(as_messydate("2012-01-01"), as_messydate("2012-02"))
is_similar(as_messydate("2012-06-02"), as_messydate("2012-02-06"))
is_similar(as_messydate("2012-06-22"), as_messydate("2012-02-06"))
```

Additionally, one can perform intersection (`md_intersect()`) and union (`md_union()`) on, 
inter alia, messy date class objects. Or 'join' that retains all elements, even if duplicated,
with `md_multiset`.

```{r set, warning=FALSE}
md_intersect(as_messydate("2012-01-01..2012-01-20"),as_messydate("2012-01"))
md_union(as_messydate("2012-01-01..2012-01-20"),as_messydate("2012-01"))
md_multiset(as_messydate("2012-01-01..2012-01-20"),as_messydate("2012-01"))
```
