---
title: "Working with messy dates"
author: "Henrique Sposito"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with messy dates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Why `{messydates}`?

Dates are often messy.
Whether historical (or ancient), future, or even recent,
we often only know approximately when an event occurred,
that it happened within a particular period,
an unreliable source means a date should be flagged as uncertain,
or sources offer multiple, competing dates.

The goal of `{messydates}` is to help with this problem by retaining and working
with various kinds of date imprecision.
\pkg{messydates} contains a set of tools for constructing and coercing into and from the 'mdate' class.
This date class implements ISO 8601-2:2019(E) and allows regular dates to be annotated to express unspecified date components, approximate or uncertain date components, date ranges, and sets of dates.

# Working with messydates: The size of the popuklation of the Roman Empire

Several historical facts are contested. Take, for example, the population sizes in the Roman Empire, which has been a topic of heated debates among some academics. Various accounts diverge in how they estimate population size and, therefore, the conclusions about the Roman population at different points in time. Below, we exemplify how \pkg{messydates} facilitate the annotation and the conversion of historical dates for analysis on the dataset of the high and low estimations for the population of mainland Italy during the Roman Empire included in the package.

```{r data, warning=FALSE}
library(messydates)
roman_population # For more information on the dataset please run `?roman_population`.
```

## Coerce to messydates

The function `as_messydate()` handles the annotation and coercion to 'mdate' class.
We can, for instance, coerce the date variable in the Roman population data to 'mdate'.

```{r messy, warning=FALSE}
roman_population$period <- as_messydate(roman_population$period)
roman_population$period
```

## Annotate

Some datasets have, for example, an arbitrary cut off point for start and end points, 
but these are often coded as precise dates when they are not necessarily the real start or end dates.
The annotate functions helps annotate uncertainty and approximation to dates.
Inaccurate start or end dates can be represented by an ".." affix
indicating "on or before", if used as a prefix,
or indicating "on or after", if used as a suffix.
Approximate dates are indicated by adding a `~` to year,
month, or day components, as well as groups of components or whole dates
to estimate values that are possibly correct.
Day, month, or year, uncertainty can be indicated by adding a `?`
to a possibly dubious date or date component.

For example, we can add these annotations to the 'mdate' variable in the Roman population data.

```{r annotate, warning=FALSE}
ifelse(roman_population$period == "-0200..-0100", on_or_before(roman_population$period), roman_population$period)
ifelse(roman_population$period == "0150..0200", on_or_after(roman_population$period), roman_population$period)
ifelse(roman_population$period == "-0100..0001", as_approximate(roman_population$period), roman_population$period)
ifelse(roman_population$period == "0001..0150", as_uncertain(roman_population$period), roman_population$period)
```

## Expand

Expand functions transform date ranges, sets of dates, and unspecified or 
approximate dates (annotated with '..', '{ , }', 'XX' or '~') into lists of dates.
As these dates may refer to several possible dates,
the function "opens" these values to include all the possible dates implied.

The ranges in the 'mdate' variable in the Roman population data can be also expanded.

```{r expand, warning=FALSE}
exp <- expand(roman_population$period) # Please run `exp` to see all the expanded dates.
lengths(exp)
```

## Contract

The `contract()` function operates as the opposite of `expand()`.
It contracts a list of dates into the abbreviated annotation of messydates.
We can contract back the 'mdate' variable in the Roman population previously expanded.

```{r contract, warning=FALSE}
tibble::tibble(contract = contract(expand(roman_population$period), collapse = FALSE))
```

## Coerce from messydates

Coercion functions coerce objects of `mdate` class to 
common date classes such as `Date`, `POSIXct`, and `POSIXlt`.
Since `mdate` objects can hold multiple individual dates,
an additional function must be passed as an argument so that multiple dates
are "resolved" into a single date.

For example, one might wish to use the earliest possible date
in any ranges of dates (`min`), the latest possible date (`max`),
some notion of a central tendency (`mean`, `median`, or `modal`),
or even a `random` selection from among the candidate dates. 
These functions are particularly useful for use with existing methods and models,
especially for checking the robustness of results.

We can coerce the 'mdate' variable in the Roman population to `Date` class.

```{r coerce, warning=FALSE}
tibble::tibble(min = as.Date(roman_population$period, min),
               max = as.Date(roman_population$period, max),
               median = as.Date(roman_population$period, median),
               mean = as.Date(roman_population$period, mean),
               modal = as.Date(roman_population$period, modal),
               random = as.Date(roman_population$period, random))
```

## Additional functionality

Several other functions are offered in the package as, for example,
we can proceed to run logical tests to the 'mdate' variable in the Roman population.
`is_messydate()` tests whether the object inherits the `mdate` class.
`is_intersecting()` tests whether there is any intersection between two messy dates.
`is_element()` similarly tests whether a messy date can be found within a messy date range or set.
`is_similar()` tests whether two dates contain similar components.
`is_precise()` tests whether certain date is precise.

```{r logical, warning=FALSE}
is_messydate(roman_population$period)
is_intersecting(as_messydate(roman_population$period[1]), as_messydate(roman_population$period[2]))
is_element(as_messydate("-0199-01-01"), as_messydate(roman_population$period[1]))
is_similar(as_messydate("-0199-01-01"), as_messydate(roman_population$period[1]))
is_precise(as_messydate(roman_population$period[1]))
```

Additionally, one can perform intersection, union, or 'join' the 'mdate' variable in the Roman population.

```{r set, warning=FALSE}
md_intersect(as_messydate(roman_population$period[1]), as_messydate("-0200-01-01"))
length(md_union(as_messydate("-0201-12-31"), as_messydate(roman_population$period[1])))
# Please run `md_union(as_messydate("-0201-12-31"), as_messydate(roman_population$period[1]))` to see all united dates.
length(md_multiset(as_messydate("-0200-01-01"), as_messydate(roman_population$period[1])))
# Please run `md_multiset(as_messydate("-0200-01-01"), as_messydate(roman_population$period[1]))` to all joined dates.
```

As well, we can add or subtract one year to the 'mdate' variable in the Roman population.

```{r operate}
roman_population$period + 1 # Adds one day to ranges
as.Date(roman_population$period, random) - 1 # Subtracts one day from a random day within range
```
