---
title: "Working with messy dates"
author: "Henrique Sposito"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with messy dates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Why `{messydates}`?

Dates are often messy.
Whether historical (or ancient), future, or even recent,
we sometimes only know approximately when an event occurred,
that it happened within a particular period,
an unreliable source means a date should be flagged as uncertain,
or different sources offer multiple, competing dates.

The goal of `{messydates}` is to help with this problem by retaining and working
with various kinds of date imprecision.
\pkg{messydates} contains a set of tools for constructing and coercing
into and from the `mdate` class.
This date class implements [ISO 8601-2_2019(E)](https://www.iso.org/standard/70908.html)
allowing regular dates to be annotated to express unspecified date components,
approximate or uncertain date components, date ranges, and sets of dates.

# Working with messydates: 2001 Battles

Take, for example, the names and dates of battles in 2001 according to [Wikipedia](https://en.wikipedia.org/wiki/List_of_battles_in_the_21st_century)
included in \pkg{messydates}.
The dates of these battles are often uncertain or approximate with different
levels of date precision being reported.

```{r data, warning=FALSE}
library(messydates)
battles <- messydates::battles
battles
```

## Coerce to messydates

Previously researchers had to remove all types of imprecision from date variables and
create multiple variables to deal with date ranges.
`{messydates}` makes it much easier to retain and work with various kinds of date imprecision.
In the 2001 battles dataset, for example, we see that dates are not consistently reported,
but `as_messydate()` still handles the coercion to `mdate` class.

```{r messy, warning=FALSE, message=FALSE}
battles$Date <- as_messydate(battles$Date)
battles$Date
```

## Annotate

The annotate functions in `{messydates}` help annotate censored, uncertain, and approximate
dates according to [ISO 8601-2_2019(E)](https://www.iso.org/standard/70908.html) standards.
Some datasets might have an arbitrary cut off point for start and end points,
that is, they are censored.
But these are often coded as precise dates when they are not necessarily
the real start or end dates.
Inaccurate start or end dates can be represented by an ".." affix
indicating "on or before", if used as a prefix,
or indicating "on or after", if used as a suffix.
In the case of the battles of 2001 dates, if we are not sure the
"Battle of Kandahar" began on the 22nd of November or, alternatively, that the
"Operation Vaksince" actually ended in the same day it began
we use `on_or_before()` and `on_or after()` to annotate these dates.

```{r censored, warning=FALSE}
battles$Date <- as_messydate(ifelse(battles$Battle == "Battle of Herat", on_or_before(battles$Date), battles$Date))
battles$Date <- as_messydate(ifelse(battles$Battle == "Operation Vaksince", on_or_after(battles$Date), battles$Date))
tibble::tibble(battles)
```

Additional annotations for approximate dates, are indicated by adding a `~` to year,
month, or day components, or whole dates to estimate values that are possibly correct.
Day, month, or year, uncertainty can also be indicated by adding a `?`
to a possibly dubious date or date component.
If we are not sure about the reliability of the sources for the
"Battle of Shawali Kowt" and we think the declared date for the battle is approximate,
we can use `as_uncertain()` or `as_approximate()` to annotate these dates.

```{r approximate, warning=FALSE}
battles$Date <- as_messydate(ifelse(battles$Battle == "Battle of Shawali Kowt", as_uncertain(battles$Date), battles$Date))
battles$Date <- as_messydate(ifelse(battles$Battle == "Battle of Sayyd Alma Kalay", as_approximate(battles$Date), battles$Date))
tibble::tibble(battles)
```

## Expand

Expand functions transform date ranges (annotated with '..'),
sets of dates (annotated with '{ , }'),
and unspecified (missing date components or annotated with 'XX'), or 
approximate dates (annotated '~') into lists of dates.
As these dates may refer to several possible dates,
the function "opens" these values to include all the possible dates implied.
Let's expand the dates in the Battles dataset.

```{r expand, warning=FALSE}
expand(battles$Date)
```

Note that to expand approximate dates one needs to declare the range to
expand approximate dates using the 'approx_range' argument in `expand()`

```{r expand_approx, eval=FALSE}
expand(battles$Date, approx_range = 1)
```

## Contract

The `contract()` function operates as the opposite of `expand()`.
It contracts a list of dates into the abbreviated annotation of messydates,
picking the most succinct representation of dates possible.
We can contract back the dates in the Battles data previously expanded.

```{r contract, warning=FALSE}
tibble::tibble(contract = contract(expand(battles$Date)))
```

## Coerce from messydates

Coercion functions coerce objects of `mdate` class to 
common date classes such as `Date`, `POSIXct`, and `POSIXlt`.
Since `mdate` objects can hold multiple individual dates,
an additional function must be passed as an argument so that multiple dates
are "resolved" into a single date.

For example, one might wish to use the earliest possible date
in any ranges of dates (`min`), the latest possible date (`max`),
some notion of a central tendency (`mean`, `median`, or `modal`),
or even a `random` selection from among the candidate dates. 
These functions are particularly useful for use with existing methods and models,
especially for checking the robustness of results.

```{r coerce, warning=FALSE}
tibble::tibble(min = as.Date(battles$Date, min),
               max = as.Date(battles$Date, max),
               median = as.Date(battles$Date, median),
               mean = as.Date(battles$Date, mean),
               modal = as.Date(battles$Date, modal),
               random = as.Date(battles$Date, random))
```

## Additional functionality

Several other functions are offered in `{messydates}`.

For example, we can run several logical tests to `mdate` variables.
`is_messydate()` tests whether the object inherits the `mdate` class.
`is_intersecting()` tests whether there is any intersection between two messy dates.
`is_element()` similarly tests whether a messy date can be found within a messy date range or set.
`is_similar()` tests whether two dates contain similar components.
`is_precise()` tests whether certain date is precise.

```{r logical, warning=FALSE}
is_messydate(battles$Date)
is_intersecting(as_messydate(battles$Date[1]), as_messydate(battles$Date[2]))
is_element(as_messydate("2001-04-17"), as_messydate(battles$Date[2]))
is_similar(as_messydate("2001-08-03"), as_messydate(battles$Date[1]))
is_precise(as_messydate(battles$Date[2]))
```

Additionally, one can perform intersection, union, or 'join' of messydates.

```{r set, warning=FALSE}
md_intersect(as_messydate(battles$Date[9]), as_messydate(battles$Date[10]))
md_union(as_messydate(battles$Date[17]), as_messydate(battles$Date[18]))
md_multiset(as_messydate(battles$Date[1]), as_messydate(battles$Date[2]))
```

As well, we can do some arithmetic operations in the `mdate` variable.

```{r operate}
tibble::tibble("one day more" = battles$Date + 1,
               "one day less" = battles$Date - "1 day")
```
